Comparison of MCMC algorithms in Stochastic Volatility Models

### 2023-08-23 writeups/research_proposal/propsal.qmd
first draft edited

## Research Goal
The objective of this research is to validate and compare the comutational methods used to estimate Bayesian Stochastic Volatiity models. Specifically, Simulation Based Calibration will be used to identify bias and errors arising from the results of computational algorithms. The source of such errors include, but are not limited to, different estimation strategies, problems with the sampler, problematic model specifications and coding mistakes. This study will compare the estimation strategies detailed in the Stochastic Volatility literature and the application of Hamiltonina Monte Carlo for estimating these class of models. Results and diagnostics from these simulations will help identify strengths and limitations that arise from these approaches and provide tools for evaluating other types models and algorithms.

@@

Kim, Sherpherd and Chib (1998) estimate the discrete time Stochastic Volatility model by proposing a simulation strategy using Bayesian methods. Specifically, they propose a method of sampling using an offset mixture normal approximation with a reweighting procedure to sample from the posterior of the SV model parameters. By linearing the model and using a mixture of Gaussian's, they were able to apply a Kalman Filter to estimate the latent volatilities of the state space models. The use of simulation and sampling are used as part of model diagnostics and evaluation. This is compared with other estimation strategies and volatility models such as autogregressive conditional heteroskedasticity (ARCH). Their specificaiton of the Stochastic Volatility model and estimation approaches provide the context for comparing the performance of different MCMC algorithms.

The simulations run by Kim, Sherpherd and Chib rely on sampling directly from conjugate priors and use of classic Markov Chain Monte Carlo algorithms, Gibbs and Metropolis Hastings. Since the paper was written, new MCMC algorithms have been developed and enabled the estimation of richer and more complicated models. Specifically, Hamiltonian Monte Carlo is an a MCMC algorithm which has been become widely available for efficiently sampling from sophisticated models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo was developed in the physics literature (Duane et al., 1987) before being applied in the statistics literature by Radford Neal through his works in Bayesean Neural Networks (Nael, 1995) and statistical computing (Neal 2011). The algorithm has since become widely available through open source developement projects such as Stan (Stan Development Team, 2017) and PyMC (2023).

New developments in MCMC algorithms and improved computing resources have enabled estimation of richer and more complicated models. However, as the development of tools to enable us to fit more complicated models, so does the need to validate the output of these tools. Cook, Gelman and Rubin (2006) introduce simulation techniques to evaluate the "correctness" of the software used to sample Bayesian models. Their contribution used CDF quantiles from the posterior samples will be uniformly distributed. Talts et al. build upon this work by developing Simulation Based Calibration to provide more robust validation of output from a MCMC sampler through the use of rank statistics.

@@

## 3.1) Stochastic Volatility Model
The univariate stochastic volatility model described by Kim, Shepherd and Chib (1998), now denoted as KSC, models the variance as a random variable following some latent stochastic process. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or "stickiness" of log volatility and the dispersion parameter $\sigma_{\eta}$ is its variance. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks.

@@

With priors for the static parameters as defined in KSC:

@@

### KSC
KSC's model uses a simulation smoother[^1] (de Jong and Shephard, 1995) to estimate from the latent states. This requires the state space model to be linear and conditionally Gaussian.

@@

Where $z_t = log(\epsilon_t^2)$ follows a log chi-squared distribution with mean -1.2704 and varaince 4.93. However, it is not simple to sample from this parameterisation of the model. KSC use a gaussian mixture model to **approximate** the first 4 moments of the log chi squared distribution. This is defined by:

@@

### Hamiltonian Monte Carlo (No-U-Turn Sampler)
Stan's default Hamiltonian Monte Carlo algorithm, the No-U-Turn Sampler, allows for direct sampling of the spcified stochastic volatility model. The HMC algorithm and more broadly the Stan programming language allows for sampling of the generative model and can flexibly handle complicated likelihood functions.

HMC will use the same priors are specified in the KSC model, however, the HMC algorithm does not require the priors to be conjugate (nor does the algorithm sample from conjugate or conditionally posterior distributions).

@@

SBC is conducted by comparing the distribution of rank statistics to the discrete uniform distribution which arises from a correctly calibrated algorithm. The procedure starts by generating datasets implied by the prior distribution and data generating process and calculating rank statistics from the resulting posterior samples. Starting with the parameters $\theta$ and data $y$, draw a sample from the prior distribution:

$$
\begin{aligned}
\theta^{sim} \sim \pi(\theta)
\end{aligned}
$$

Generate a dataset given by the prior draws and data generating process p.

$$
\begin{aligned}
y^{sim} \sim p(y|\theta^{sim})
\end{aligned}
$$

Then take draws from the posterior distribution generated by a MCMC algorithm or estimation strategy (i.e estimate the model on the simulated data using either KSC's strategy or HMC):
$$
\begin{aligned}
\{\theta_1,\dots , \theta_{L}\} \sim p(\theta | y^{sim})
\end{aligned}
$$

@@

That is, the posterior averaged over the joint distribution follows the same distribution as the prior. The procedure of generating posterior samples implicitly performs this integral. The expression on the right hand side is proportional to the prior density and the simulation if done correctly should produce samples from the prior distribution. Therefore, any deviation of the posterior samples from the prior distribution suggests an error - either in computation, code or overall analysis.

@@

[^2]: Proof of this result in Talts et al.

@@

This completes one iteration of SBC. Multiple iterations are run and the rank statistics are calculated for each parameter. The resulting rank statistics are compared to the discrete uniform distribution to determine if any problematic features or errors exist.

@@

# Results
SBC for the HMC algorithm and the KSC Gaussian approxmiation are conducted with simulated dataset sizes of 1000 observations, 1000 simulation interations with post warmup/burnin samples of 999 (which gives 1000 possible rank statistics). The histogram is drawn with 20 bins and 50 observations in each bin which is the expected count for a discrete uniform distribution. The preliminary results focus on a subset of parameters since there are parameters for each latent state (i.e one for each data observation) plus the static parameters.

Figure 1 shows the the distribution of rank statistics for the HMC algorithm, with the horizontal black line at 50 representing the discrete uniform distribution. Phi and sigma_sqd look relatively uniform. However mu and h.995 have some lumpiness which may suggest a lack of calibration.

@@

One of the drackbacks of SBC is that any deviation from uniform from a calibrated analysis maybe due to finite samples. Increasing the number of iterations from 1000 to 5000 shows more consistent uniform behaviour from all parameters (Figure 2). This suggests that the HMC algorithm and overall analysis is well calibrated.

@@

Figure 3 displays the results from KSC's Gaussian approximation. Compared to HMC, the distribution of ranks deviate heavily from the expected discrete uniform distribution. The lack of calibration relative to the HMC algorithm is expected. The Gaussian Mixture Approximiation only approximates the log chi squared distribution. The HMC strategy samples directly from the same likelihood as the data generating process.

@@

Figure 4 and 5 are the chi squared statistic histograms results of the HMC algorithm and KSC algorithm respectively. The vertical black line marks the 95th quantile which represents the critical value for hypothesis testing. The HMC chi squared statistics look like they are sampled from the correct distribution whereas KSC are inconsistent with the expected distribution.

@@

# Conclusion and next steps
KSC also also apply a Metropolis Hastings reweighting to correct the approximation of the error distribution. According to the authors, the correction ensures that samples are from the exact posterior density. The next step is to perform this correction and evaluate the SBC results.

Another factor that may affect sampling performance is the parameterisation of the model. This can cause significant problems to samplers in models with increasing complexity. A classic example of this is Neal's funnel which arises from Hierarchical models - causing bias and performance issues with the HMC sampler. The solution is to rewrite the model with non centered parameterisation which significantly improves the performance of the sampler. Another step in this research is to consider the results of SBC under different model parameterisations. The models and parametiersations for this research are summarised in the below table:

@@

@@

# Appendix A: Mixture Gaussian weights
Weights

>>>>

## Research Goal
The objective of this research is to validate and compare the comutational methods used to estimate Bayesian Stochastic Volatiity models. Specifically, Simulation Based Calibration will be used to identify biases and errors arising from the results of Markov Chain Monte Carlo (MCMC) algorithms. The source of any miscalibration include, but are not limited to, estimation strategies, choice of sampler, model specifications and coding mistakes. This study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo for estimating these class of models. Results and diagnostics from these simulations will help identify strengths and limitations that arise from these approaches.

<!-- and provide tools for evaluating other types models and algorithms.  -->

@@

Kim, Sherpherd and Chib (1998) estimate the discrete time stochastic volatility model by proposing a simulation strategy using Bayesian methods. Specifically, they propose a method of sampling using an offset mixture normal approximation with a reweighting procedure to sample from the posterior of the stochastic volatility model parameters. By linearing the model and using a mixture of Gaussian's, they were able to apply a Kalman Filter to estimate the posterior density over the latent volatilities and static parameters. The use of simulation and sampling are used as part of model diagnostics and evaluation. This is compared with other estimation strategies and volatility models such as autogregressive conditional heteroskedasticity (ARCH). Their specificaiton of the stochastic volatility model and estimation approaches provide the context for comparing the performance of different MCMC algorithms.

The simulations run by Kim, Sherpherd and Chib rely on sampling directly from conjugate priors and use of classic MCMC algorithms, Gibbs and Metropolis Hastings. Since the paper was written, new MCMC algorithms have been developed and enabled the estimation of richer and more complicated models. Specifically, Hamiltonian Monte Carlo is an a MCMC algorithm which has been become widely available for efficiently sampling from sophisticated models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo was developed in the physics literature (Duane et al., 1987) before being applied in the statistics literature by Radford Neal through his works in Bayesean Neural Networks (Nael, 1995) and statistical computing (Neal 2011). The algorithm has since become widely available through open source developement projects such as Stan (Stan Development Team, 2023) and PyMC (2023).

The key innovation of Hamiltonian Monte Carlo is using the gradients of the target posterior distribution to generate an efficient path for the sampler to explore. Unlike Random Walk Metropolis Hastings, it takes advantage of the geometry of the posterior to determine its next step. A comprehensive explanation of the sampler is beyond the scope of this research and can be found in the above references.

New developments in MCMC algorithms and improved computing resources have enabled estimation of richer and more complicated models. However, as the development of tools to enable us to fit more complicated models, so does the need to validate the output of these tools. Cook, Gelman and Rubin (2006) introduce simulation techniques to evaluate the "correctness" of the software used to sample Bayesian models. Their contribution identified that cumulative density quantiles from the posterior samples will be uniformly distributed. Talts et al. build upon this work by developing Simulation Based Calibration to provide more robust validation of output from a MCMC sampler through the use of rank statistics.

@@

## 3.1) Stochastic Volatility Model
The univariate stochastic volatility model described by Kim, Shepherd and Chib (1998), models the variance as a random variable following some latent stochastic process. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or "stickiness" of log volatility and the dispersion parameter $\sigma_{\eta}$ is its variance. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks.

@@

Priors for the static parameters are defined below with $\mu$ and $\sigma^2$ being conjugate priors:

@@

### Gaussian Mixture Approximation
Kim Shepherd and Chib estimate the stochastic volatility model using a Kalman Filter and smoother[^1] (de Jong and Shephard, 1995) to compute the posterior distribution over the latent states. This requires the state space model to be linear and conditionally Gaussian.

@@

Where $z_t = log(\epsilon_t^2)$ follows a log chi-squared distribution with mean -1.2704 and varaince 4.93. However, it is not simple to sample from this parameterisation of the model. Kim, Sherpherd and Chib use a mixture of Gaussians to **approximate** the first 4 moments of the log chi squared distribution. This is defined by:

@@

### Hamiltonian Monte Carlo (No-U-Turn Sampler)
The Stan programming language's implementation of Hamiltonian Monte Carlo will be used for this study. Stan's default algorithm, the No-U-Turn Sampler (Hoffman & Gelman 2014), allows for direct sampling of the spcified stochastic volatility model. Hamiltonian Monte Carlo allows for sampling of the generative model and can flexibly handle complicated likelihood functions. This approach will use the same priors are specified in the Gaussian mixture approximation. However, Hamiltonian does not sample directly from the analytical form of conjugate or conditionally conjugate distribution, so different priors can be chosen if necessary.

@@

SBC is conducted by comparing the distribution of rank statistics to the discrete uniform distribution which arises when an algorithm is correctly calibrated. The procedure starts by taking draws from the prior distribution and creating datasets implied by each draw. Rank statistics are then calculated on the posterior samples estimated from the simulated data. Let $\theta$ be a parameter and $y$ represent the dataset. Start with draw a sample from the prior distribution:
$$
\begin{aligned}
\theta^{sim} \sim \pi(\theta)
\end{aligned}
$$

Generate a dataset given by the prior draws.

$$
\begin{aligned}
y^{sim} \sim \pi (y|\theta^{sim})
\end{aligned}
$$

Then take draws from the posterior distribution generated by a MCMC algorithm or estimation strategy (Hamiltonian Monte Carlo or Gaussian apprimation in our case).

$$
\begin{aligned}
\{\theta_1,\dots , \theta_{L}\} \sim \pi (\theta | y^{sim})
\end{aligned}
$$

@@

That is, the posterior averaged over the joint distribution follows the same distribution as the prior. The procedure of generating posterior samples implicitly performs this integral since the expression on the right of the integral is proportional to the prior density. Therefore, any deviation of the posterior samples from the prior distribution suggests an error - either in computation, code or overall analysis.

<!-- The simulation if done correctly should produce samples from the prior distribution.  -->

@@

[^2]: Proof of this result in Talts et al 2018.

@@

This completes one iteration of SBC. To complete the algorithm, multiple iterations are run and the rank statistics are calculated for each parameter. The resulting rank statistics are compared to the discrete uniform distribution to determine if any problematic features or errors exist.

@@

# Results
SBC for Hamiltonian Monte Carlo and the Gaussian approxmiation are performed with simulated dataset sizes of 1000 simulation interations, 1000 observations, and post warmup/burnin samples of 999 (which gives 1000 possible rank statistics). The histogram is drawn with 20 bins and 50 observations in each bin which is the expected count for a discrete uniform distribution. This algorithm is summarised below:

| **1) for n in 1000 iterations:**
|   2) Draw from prior: $\theta^{sim}\sim\pi (\theta)$
|   3) Simulate dataset with 1000 observations: $y^{sim} \sim p(y|\theta^{sim})$
|   4) Draw 999 posterior samples (post warmup) $\{\theta_1,\dots , \theta_{L}\} \sim p(\theta | y^{sim})$
|   5) Compute rank statistics $r = rank(\{\theta_1,\dots , \theta_{L}\}, \theta^{sim})$

Figure 1 shows the the distribution of rank statistics for Hamiltonian Monte Carlo, with the horizontal black line at 50 representing the discrete uniform distribution. The preliminary figures focus on the static parameters and arbitrarily chosen states (since there is one state parameter per data observation). Phi and sigma_sqd look relatively uniform. However mu and h.995 have some lumpiness which may suggest a lack of calibration.

@@

One of the drackbacks of SBC is that any deviation from uniform from a calibrated analysis maybe due to finite samples. Figure 2 shows more consistent uniform bahviour from all parameters after increasing the number of iterations from 1000 to 5000. This suggests that Hamiltonian Monte Carlo and overall analysis is well calibrated.

@@

Figure 3 shows results for the Gaussian approximation. The distribution of the rank statistics deviate heavily from the expected discrete uniform distribution, with various peaks at each end of the histogram as well as peaks in the middle. This lack of calibration is expected. The Gaussian approximiation only approximates the log chi squared distribution, where as Hamiltonian Monte Carlo samples directly from the same likelihood as the data generating process.

@@

Figure 4 are the chi squared statistic histograms results of Hamiltonian Monte Carlo and Gaussian approximation respectively. The vertical black line marks the 95th quantile which represents the critical value for hypothesis testing. The HMC chi squared statistics look like they are sampled from the correct distribution, with 6.2\% of the samples greater than the 95\% critical value (doing this calculation for 5000 iterations leads to 4.8\% of samples in the critical area). Figure 5 shows the majority of chi squared statistics for Gaussian approximation within the critical area, suggesting the estimation strategy is not well calibrated.

@@

# Conclusion and next steps
There are two key extensions to the simulation results worth exploring.

**1) Metropolis Hastings reweighting for KSC model**

Kim, Sherpherd and Chib apply a Metropolis Hastings reweighting to the posterior samples to correct for the approximation of the error distribution. According to the authors, the correction ensures that samples are from the exact posterior density. The next step is to perform this correction and evaluate the SBC results. I'd expect the Gaussian mixture approximation diagnostics to improve after this correction.

**2) Centered and non centered parameterisation**

The parameterisation of a model can affect the performance of a sampler. This can cause significant problems in getting unbiased MCMC samples for complex models. An example of this is Neal's funnel (Neal 2003) which arises in hierarchical models which causes performance issues with the Hamiltonian Monte Carlo sampler. Rewriting the model with non centered parameterisation significantly improves the performance of MCMC. Another consideration for this research is to consider both centered and non centered parametersations of the stochastic volatility model to see if there are any improvements to the simulation diagnostics.

The models and parametiersations for this research are summarised in the below table:

@@

## Plan
Week 7: Reflect and integrate feedback from presentation into plan for the rest of the semester

Week 8: Finish off any simulation or modelling not yet complete (listed in above matrix)

Week 9: Write up any other results and have a first draft ready for feedback

Week 10-11: Action any feedback, prepare for final submission and presentation

Week 12: Submit

@@

Abril-Pla O, Andreani V, Carroll C, Dong L, Fonnesbeck CJ, Kochurov M, Kumar R, Lao J, Luhmann CC, Martin OA, Osthege M, Vieira R, Wiecki T, Zinkov R. (2023) PyMC: a modern, and comprehensive probabilistic programming framework in Python.

Duane, Simon; Kennedy, Anthony D.; Pendleton, Brian J.; Roweth, Duncan (1987). "Hybrid Monte Carlo". Physics Letters B. 195 (2): 216–222.

Hoffman, M. D., & Gelman, A. (2014). The No-U-Turn sampler: adaptively setting path lengths in Hamiltonian Monte Carlo. J. Mach. Learn. Res., 15(1), 1593-1623.

Kim, Sangjoon, Neil Shephard, and Siddhartha Chib. 1998. “Stochastic Volatility: Likelihood Inference and Comparison with ARCH Models.” Review of Economic Studies 65: 361–93.

Neal, R.M. (1995) Bayesian Learning for Neural Networks. Ph.D. Thesis, Graduate Department of Computer Science, University of Toronto, Toronto.

Neal, R. M. (2003). Slice Sampling. The Annals of Statistics, 31(3), 705–741. http://www.jstor.org/stable/3448413

Neal, R.M. (2011) MCMC Using Hamiltonian Dynamics. In Handbook of Markov Chain Mone Carlo, CRC Press, New York

Stan Development Team. 2023. Stan Modeling Language Users Guide and Reference Manual, 2.32. https://mc-stan.org

Talts, S., Betancourt, M., Simpson, D., Vehtari, A., & Gelman, A. (2018). Validating Bayesian inference algorithms with simulation-based calibration. arXiv preprint arXiv:1804.06788.

# Appendix A: Mixture Gaussian weights
